---
title: "Reproducible Report on Covid19 Data"
author: "D. Fernandez"
date: "04/23/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

<p>This is an analysis of the COVID19 data taken from the beginning of the pandemic until present day. This data was obtained from the John Hopkins [github](https://github.com/CSSEGISandData/COVID-19) site. We will observe the number of COVID cases and deaths throughout the United States, observe only "new" cases and deaths, and look at a linear model that looks at the deaths per 1000 as a function of cases per 1000. </p> 
<p>
Let us first set up our libraries and import our data. We will then tidy our data and create our graphs.
</p>

```{r, include = TRUE, message = FALSE, warning = FALSE}
#Set up the libraries
library(tidyverse)
library(lubridate)
library(scales)
library(ggplot2)

# Get data from the john hopkins github
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_global.csv", "time_series_covid19_confirmed_US.csv", "time_series_covid19_deaths_US.csv")
urls <- str_c(url_in,file_names)
```

```{r, include = TRUE, message = FALSE, warning = FALSE}
#Take CSVs and assign them to the proper variable
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
US_cases <- read_csv(urls[3])
US_deaths <- read_csv(urls[4])
```

```{r, include = TRUE, message = FALSE, warning = FALSE}
#Tidy the variable global_cases
global_cases <- global_cases %>% 
  pivot_longer(cols = -c('Province/State','Country/Region',Lat,Long), names_to = "date", values_to = "cases") %>%
  select(-c(Lat,Long))

#Tidy the variable 
global_deaths <- global_deaths %>% 
  pivot_longer(cols = -c('Province/State','Country/Region',Lat,Long), names_to = "date", values_to = "deaths") %>%
  select(-c(Lat,Long))

#Combine global_cases with global_deaths to make a single global variable
global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = 'Country/Region',
         Province_State = 'Province/State') %>%
  mutate(date = mdy(date))

#Filter results where cases are positive
global <- global %>% filter(cases > 0)

#Tidy variable US_cases
US_cases <- US_cases %>% 
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

#Tidy variable US_deaths
US_deaths <- US_deaths %>% 
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

#Combine US_cases and US_deaths to make single US variable
US <- US_cases %>%
  full_join(US_deaths)

#Combine Province_State and Country_Region in global
global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)

#Get population data 
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

#Combine global variable with population data
global <- global %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)

#Create variable US_by_state
US_by_state <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

#Create US_totals variable
US_totals <- US_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()
```

```{r, include = TRUE, message = FALSE, warning = FALSE}
#Create graph that shows the number of COVID19 cases and deaths in the US
US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID 19 in US", y = NULL)
```

#### Observations
<p>
We can see in our graphs that the total number of COVID19 cases and deaths rose sharply in the early part of 2020. The rate of growth slows in July of 2020 and levels off at the beginning of 2021. We can also observe that the rate of cases and deaths share a close relationship (as cases increase, so do deaths). Next, let's examine one state and see if there are any differences in the graph.
</p>
```{r, include = TRUE, message = FALSE, warning = FALSE}
#Create graph that shows the number of COVID19 cases and deaths in the state of New York
state <- "New York"
US_by_state %>%
  filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID 19 in New York", y = NULL)
```

#### Observations
<p>
We can see that our graph for the state of New York shares a close resemblance to our US graph. Next, let us create a new graph which will examine the number of new COVID19 cases and deaths, rather than the total number.
</p>
```{r, include = TRUE, message = FALSE, warning = FALSE}
#Add new variables US_by_state and US_totals
US_by_state <- US_by_state %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
US_totals <- US_totals %>%
  mutate(new_cases = cases -lag(cases),
         new_deaths =deaths - lag(deaths))

#Create graph for the US_totals which will show new cases and new deaths
US_totals %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) + scale_y_log10() +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US", y = NULL)
```

#### Observations
<p>
Like the previous graphs, we see a sharp uptick in the number of COVID19 cases/deaths at the beginning of 2020. However, as time goes on we can see that the number of new cases/death waxes and wanes over time. There are many possible explanations as to why we see these fluctuations. Over the course of the pandemic there have been different guidelines given by the CDC overtime. Also different states had mask mandates while others did not. What's more is that different strains of the virus (some more contagious) have popped up and could be responsible for the increase.
</p>
<p>
Next we will take a look at a linear model that examines the deaths per 1000 as a function of cases per 1000.
</p>
```{r, include = TRUE, message = FALSE, warning = FALSE}
#Create new variable US_state_totals by taking US_by_state and adding cases_per_thou and deaths_per_thou
US_state_totals <- US_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases), population = max(Population), cases_per_thou = 1000* cases / population, deaths_per_thou = 1000* deaths / population) %>%
  filter(cases > 0, population > 0)

#Create a linear model that looks at the deaths per 1000 as a function of cases per 1000
mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)

US_state_totals %>% slice_max(cases_per_thou)
x_grid <- seq(1,151)
new_df <- tibble(cases_per_thou = x_grid)
US_state_totals %>% mutate(pred = predict(mod))
US_tot_w_pred <- US_state_totals %>% mutate(pred = predict(mod))

US_tot_w_pred %>% ggplot() + geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")
  
```

#### Observations
For this model we have our actual cases in blue and our predictor variable is in red. The model does an adequate job of predicting at the lower and higher ends of the graph, but we can see that there is a lot of variance in the middle. Perhaps there are other variables in play that are influencing our model.


## Bias
When it comes to this dataset and analysis, there are places where there may be some bias. Bias is a distortion which results in information not being a true representation of the reality we are trying to investigate. Unfortunately, the COVID19 pandemic has become politicized. Some opinionated political pundits have sowed distrust and doubt into our medical institutions. This has lead people to avoiding the COVID19 vaccine and instead use unproven/ineffective remedies. I have a strong suspicion that the actual number of cases is higher than the recorded cases due to these factors. Also, there is a possibility that the recorded number of deaths is higher than the actual number. If a terminal cancer patient contracts COVID19 a week before they die, does that count as a COVID19 death? Situations like this might inflate the recorded number of deaths. However, this is just speculation.



